---
title: "HW5 - Make Data Move"
author: "Pernille Dahl"
date: "FIND KORREKT DATO"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
Today, you will load a filtered gapminder dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks below.**

## Get the necessary packages
First, start with installing the relevant packages 'tidyverse', 'gganimate', and 'gapminder'.

```{r libraries, echo = FALSE}
install.packages("gifski")
install.packages("av")
install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

We see an interesting spread with an outlier to the right. Answer the following questions, please:


1. *Why does it make sense to have a log10 scale on x axis?*
It makes sence to have a log10 scale on the x axis, because is allows you to visualize the differences in GDP across countries and making a linear scale can make it defficult to see patterns or differences in the data. 

2. *Who is the outlier (the richest country in 1952 - far right on x axis)?* 
To find out who the outlier is I create a new dataframe, where I extract the data from 1952 and then use the top_n-function to select the country with the hights GDP:

```{r}
data_1952 <- gapminder %>% filter(year == 1952)

data_1952 %>% 
  arrange(desc(gdpPercap)) %>%
  top_n(1)
```
With this solution, the correct answer Kuwait.


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

Tasks:

3. *Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation", which you might want to eliminate)*
```{r}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  scale_x_log10() 
```
Here I added "colour = continent" to make the countries in different colours. 
```{r}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10() +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "GDP per Capita (in dollars)", y = "Life Expectancy (in years)", color = "Continent") +
  theme(legend.title = element_text(size = 12), 
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) 
```
When I use the scale_x_continuous() and labels with scales::comma I eliminate scientific notation. 
And with the element_text(size = ) I adjust axis labels and units.
Here I therefore changed it to regular notation with commas, which makes it more readable.

4. *What are the five richest countries in the world in 2007?*

In order to find the five richest countries in the world in 2007 I want to use dplyr and its arrange-function and select the top_n rows:

```{r}
library(dplyr)

top_5_richest_countries_2007 <- gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap)) %>%
  head(5)

top_5_richest_countries_2007
```
The five richest countries in 2007 is therefore Norway, Kuwait, Singapore, United States and Ireland.

##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smoothes the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 


5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*
I have chosen to work only with the last animation.
To have the title change in sync with the animation I can use the transition-time function to specify the variable that the title should transition with.
```{r}
anim2 + labs(title = "{closest_state}") +
  transition_time(year) +
  shadow_mark()
```
Here's to use tranision-states for the same purpose:
```{r}
anim2 + labs(title = "{closest_state}") +
  transition_states(year, transition_length = 2, state_length = 1) +
  shadow_mark()
```

6. *Can you made the axes' labels and units more readable? Consider expanding the abreviated lables as well as the scientific notation in the legend and x axis to whole numbers.*
I want to use the labs() function to change the labels and units on the axes:
```{r}
anim2 + 
  labs(x = "GDP per capita", y = "Life Expectancy", size = "Population", title = "GDP and Life Expectancy by Year") +
  scale_x_continuous(label = function(x) format(x, scientific = FALSE)) +
  scale_size_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )
```
Here I have set the x axis lab to GDP per captia, the y axis to Life Expectancy, size label to Population an the title to GDP and Life Expectancy by Year.

7. *Come up with a question you want to answer using the gapminder data and write it down. Then, create a data visualisation that answers the question and explain how your visualization answers the question. (Example: you wish to see what was mean life expectancy across the continents in the year you were born versus your parents' birth years). [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset and download more at https://www.gapminder.org/data/ ]*
I have chosen the following question:

#How has the GDP per capita and life expectancy of countries in Africa changed over time? 

To answer my question i create a line chart with years on the x axis, GDP per capita on the y axis and the lines colored by continent. 

Now I create my line chart:
```{r}
ggplot(subset(gapminder, continent == "Africa"), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10() 
```
Here I filtered the chart only to show data for African countries. 

#Saving

